Toronto’s Shelter System Under Pressure: Investigating Homelessness and Bed Occupancy

### 2025-06-18 paper/paper.qmd
changed paper to python

```{r setup, include=FALSE}

library(ggplot2)
library(dplyr)
library(lubridate)
library(here)
library(knitr)

data <- read.csv(here("data/analysis_data/analysis_data.csv"))

@@

```{r, warning=FALSE}
#| label: tbl-cleaned_data
#| echo: false
data_head <- head(data)

data_subset <- data_head[, c("occupancy_date", "service_user_count", "program_model")]

kable(data_subset, col.names = c('Occupancy Date', 'Service User Count', 'Program Model'), 
      align = "ccc", caption = "Sample of Cleaned Shelter Occupancy Data")

@@

```{r, warning=FALSE}
#| label: fig-service-user-trend
#| fig-cap: Distribution of Service User Count Over Time by Date of Occupancy
#| echo: false

data$occupancy_date <- as.Date(data$occupancy_date, format="%Y-%m-%d")

ggplot(data, aes(x = occupancy_date, y = service_user_count)) +
  geom_jitter(color = "gray", alpha = 0.5, width = 0.2, height = 0.2, size = 0.1) + # Adjust size here
  theme_minimal() +
  labs(x = "Date of Sample Collection", y = "Service User Count") +
  scale_y_continuous(limits = c(0, 100)) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y")

@@

```{r, warning=FALSE}
#| label: tbl-observation-by-year
#| echo: false

observations_by_year <- table(data$year)

observations_table <- data.frame(
  Year = names(observations_by_year),
  Number_of_Observations = as.vector(observations_by_year)
)

mean_occupancy_rate <- mean(data$occupancy_rate_beds, na.rm = TRUE)
sd_occupancy_rate <- sd(data$occupancy_rate_beds, na.rm = TRUE)

mean_service_user_count <- mean(data$service_user_count, na.rm = TRUE)
sd_service_user_count <- sd(data$service_user_count, na.rm = TRUE)

mean_capacity_actual_bed <- mean(data$capacity_actual_bed, na.rm = TRUE)
sd_capacity_actual_bed <- sd(data$capacity_actual_bed, na.rm = TRUE)

kable(observations_table, caption = "Number of observations by year")
```

```{r, warning=FALSE}
#| label: tbl-summary-statistics
#| echo: false
#| 
summary_stats <- data.frame(
  Variable = c("Occupancy Rate (Beds)", "Service User Count", "Capacity (Actual Beds)"),
  Mean = c(mean_occupancy_rate, mean_service_user_count, mean_capacity_actual_bed),
  SD = c(sd_occupancy_rate, sd_service_user_count, sd_capacity_actual_bed)
)

kable(summary_stats, caption = "Mean and standard deviation of key variables")

@@

```{r, warning=FALSE}
#| label: fig-occupancy-rate
#| fig-cap: Average Occupancy Rate of Beds by year
#| echo: false

yearly_data <- data %>%
  group_by(year) %>%
  summarise(Avg_Occupancy_Rate_Beds = mean(as.numeric(occupancy_rate_beds), na.rm = TRUE), 
            .groups = "keep")

ggplot(yearly_data, aes(x = factor(year), y = Avg_Occupancy_Rate_Beds)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_text(aes(label = round(Avg_Occupancy_Rate_Beds, 1)), vjust = -0.5, color = "black") +
  labs(x = "Year", y = "Average Occupancy Rate (Beds)") +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

@@

```{r, warning=FALSE}
#| label: fig-bed-capacity
#| fig-cap: Actual Bed Capacity by Program Model and year
#| echo: false
filtered_data <- data %>%
  filter(program_model %in% c("Emergency", "Transitional"))

yearly_bed_capacity <- filtered_data %>%
  group_by(year, program_model) %>%
  summarise(Total_Bed_Capacity = sum(capacity_actual_bed, na.rm = TRUE), 
            .groups = "keep")

ggplot(yearly_bed_capacity, aes(x = factor(year), y = Total_Bed_Capacity, fill = program_model)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Year", y = "Actual Bed Capacity") +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal()

@@

```{r , warning=FALSE}
#| label: fig-service-user-count
#| fig-cap: Service User Count Over Months by year
#| echo: false
monthly_data <- data %>%
  group_by(year, month) %>%
  summarise(service_user_count = sum(service_user_count, na.rm = TRUE), 
            .groups = "keep")

ggplot(monthly_data, aes(x = month, y = service_user_count, color = factor(year), group = year)) +
  geom_line(size = 1) +
  labs(x = "Month", y = "Service User Count", color = "Year") +
  theme_minimal() +
  scale_x_discrete(limits = month.abb) +
  scale_y_continuous(labels = scales::comma)

@@

```{r, warning=FALSE}
#| label: fig-service-sector
#| fig-cap: Service User Count by sector for Each year
#| echo: false
library(scales)

sector_data <- data %>%
  group_by(year, sector) %>%
  summarise(Total_Service_User_Count = sum(service_user_count, na.rm = TRUE), .groups = "drop")

sector_data_filtered <- sector_data %>%
  filter(sector != "Families")

ggplot(sector_data_filtered, aes(x = sector, y = Total_Service_User_Count, fill = sector)) +
  geom_bar(stat = "identity") +
  labs(x = "Sector", y = "Total Service User Count") +
  facet_wrap(~ year, ncol = 2) +
  scale_y_continuous(labels = comma) +  # Format y-axis with whole numbers
  scale_fill_discrete(name = "Sector") +  # Change legend title
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  coord_cartesian(ylim = c(0, NA))

>>>>

```{python setup, include=FALSE}
#| include: false

import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
from datetime import datetime
from matplotlib.ticker import FuncFormatter

data = pd.read_csv("../data/analysis_data/analysis_data.csv")

@@

```{python, warning=FALSE}
#| label: tbl-cleaned_data
#| tbl-cap: Sample of Cleaned Shelter Occupancy Data
#| echo: false

# Subset and preview
data_subset = data.loc[:, ["occupancy_date", "service_user_count", "program_model"]].head()

# Rename columns for display
data_subset.columns = ["Occupancy Date", "Service User Count", "Program Model"]

data_subset

@@

```{python, warning=FALSE}
#| label: fig-service-user-trend
#| fig-cap: Distribution of Service User Count Over Time by Date of Occupancy
#| echo: false

#prepare data
data["occupancy_date"] = pd.to_datetime(data["occupancy_date"])

# Apply jitter manually
x = data["occupancy_date"]
y = data["service_user_count"]
x_jittered = x + pd.to_timedelta(np.random.uniform(-0.2, 0.2, size=len(x)), unit="D")
y_jittered = y + np.random.uniform(-0.2, 0.2, size=len(y))

# Plot
plt.figure(figsize=(10, 5))
plt.scatter(x_jittered, y_jittered, color="gray", alpha=0.5, s=1)

# Axis formatting
plt.ylim(0, 100)
plt.xlabel("Date of Sample Collection")
plt.ylabel("Service User Count")
plt.xticks(rotation=45)

# Set yearly x-axis ticks
years = pd.date_range(start=x.min(), end=x.max(), freq="YS")
plt.xticks(years, [d.strftime("%Y") for d in years])

plt.grid(False)
plt.tight_layout()
plt.show()

@@

```{python, warning=FALSE}
#| label: tbl-observation-by-year
#| tbl-cap: Number of observations by year
#| echo: false
data["year"] = pd.to_datetime(data["occupancy_date"]).dt.year

# Create summary table
observations_table = (
    data["year"]
    .value_counts()
    .sort_index()
    .reset_index()
    .rename(columns={"index": "Year", "year": "Number of Observations"})
)

# Display summary table
observations_table
```

```{python, warning=FALSE}
#| label: tbl-summary-statistics
#| tbl-cap: Mean and standard deviation of key variables
#| echo: false

# Compute summary stats
summary_stats = pd.DataFrame({
    "Variable": [
        "Occupancy Rate (Beds)",
        "Service User Count",
        "Capacity (Actual Beds)"
    ],
    "Mean": [
        data["occupancy_rate_beds"].mean(),
        data["service_user_count"].mean(),
        data["capacity_actual_bed"].mean()
    ],
    "SD": [
        data["occupancy_rate_beds"].std(),
        data["service_user_count"].std(),
        data["capacity_actual_bed"].std()
    ]
})

# Round for cleaner display (optional)
summary_stats = summary_stats.round(5)

# Display the table
summary_stats

@@

```{python, warning=FALSE}
#| label: fig-occupancy-rate
#| fig-cap: Average Occupancy Rate of Beds by year
#| echo: false

# Group and compute average occupancy rate
yearly_data = (
    data.groupby("year")["occupancy_rate_beds"]
    .mean()
    .reset_index(name="Avg_Occupancy_Rate_Beds")
)

# Plot bar chart
plt.figure(figsize=(8, 5))
bars = plt.bar(
    yearly_data["year"].astype(str),
    yearly_data["Avg_Occupancy_Rate_Beds"],
    color="skyblue"
)

# Add labels above bars
for bar in bars:
    height = bar.get_height()
    plt.text(bar.get_x() + bar.get_width() / 2, height + 1, f"{height:.1f}",
             ha="center", va="bottom", color="black")

# Axis labels and formatting
plt.xlabel("Year")
plt.ylabel("Average Occupancy Rate (Beds)")
plt.ylim(0, yearly_data["Avg_Occupancy_Rate_Beds"].max() * 1.1)
plt.tight_layout()
plt.grid(False)
plt.show()
```

@@

```{python, warning=FALSE}
#| label: fig-bed-capacity
#| fig-cap: Actual Bed Capacity by Program Model and year
#| echo: false

# Filter for Emergency and Transitional only
filtered_data = data[data["program_model"].isin(["Emergency", "Transitional"])]

# Group and summarize bed capacity by year and program model
yearly_bed_capacity = (
    filtered_data
    .groupby(["year", "program_model"])["capacity_actual_bed"]
    .sum()
    .reset_index(name="Total_Bed_Capacity")
)

# Prepare bar positions for grouped bars
years = sorted(yearly_bed_capacity["year"].unique())
models = ["Emergency", "Transitional"]
bar_width = 0.35
x = range(len(years))

# Initialize plot
fig, ax = plt.subplots(figsize=(8, 5))

# Plot each program model as separate bars
for i, model in enumerate(models):
    subset = yearly_bed_capacity[yearly_bed_capacity["program_model"] == model]
    offsets = [pos + (i - 0.5) * bar_width for pos in x]
    ax.bar(offsets, subset["Total_Bed_Capacity"], width=bar_width, label=model)

# Axes formatting
ax.set_xticks(x)
ax.set_xticklabels([str(y) for y in years])
ax.set_xlabel("Year")
ax.set_ylabel("Actual Bed Capacity")
ax.legend(title="Program Model")
ax.yaxis.set_major_formatter(FuncFormatter(lambda x, _: f"{int(x):,}"))  # Add commas
plt.tight_layout()
plt.grid(False)
plt.show()

@@

```{python, warning=FALSE}
#| label: fig-service-user-count
#| fig-cap: Service User Count Over Months by year
#| echo: false

# Ensure occupancy_date is datetime and year/month columns exist
data["occupancy_date"] = pd.to_datetime(data["occupancy_date"])
data["year"] = data["occupancy_date"].dt.year
data["month"] = data["occupancy_date"].dt.strftime("%b")  # e.g., Jan, Feb

# Group by year and month, summing service_user_count
monthly_data = (
    data.groupby(["year", "month"])["service_user_count"]
    .sum()
    .reset_index()
)

# Set month order
month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
monthly_data["month"] = pd.Categorical(monthly_data["month"], categories=month_order, ordered=True)
monthly_data = monthly_data.sort_values(["year", "month"])

# Plot
plt.figure(figsize=(10, 5))
for year, group in monthly_data.groupby("year"):
    plt.plot(group["month"], group["service_user_count"], label=year, linewidth=1)

plt.xlabel("Month")
plt.ylabel("Service User Count")
plt.legend(title="Year")
plt.grid(False)
plt.tight_layout()
plt.gca().yaxis.set_major_formatter(FuncFormatter(lambda x, _: f"{int(x):,}"))  # Comma formatting
plt.show()

@@

```{python, warning=FALSE}
#| label: fig-service-sector
#| fig-cap: Service User Count by sector for Each year
#| echo: false

import pandas as pd
import matplotlib.pyplot as plt
from matplotlib.ticker import FuncFormatter

# Group data
sector_data = (
    data.groupby(["year", "sector"])["service_user_count"]
    .sum()
    .reset_index(name="Total_Service_User_Count")
)

# Filter out 'Families' if needed
sector_data = sector_data[sector_data["sector"] != "Families"]

# Consistent color map
sector_colors = {
    "Men": "#F8766D",
    "Mixed Adult": "#7CAE00",
    "Women": "#00BFC4",
    "Youth": "#C77CFF"
}

# Define layout
years = sorted(sector_data["year"].unique())
ncols = 2
nrows = -(-len(years) // ncols)

fig, axes = plt.subplots(nrows=nrows, ncols=ncols, figsize=(12, 4 * nrows), constrained_layout=True)
axes = axes.flatten()

# Plot each year
for i, year in enumerate(years):
    ax = axes[i]
    subset = sector_data[sector_data["year"] == year]
    ax.bar(
        subset["sector"],
        subset["Total_Service_User_Count"],
        color=[sector_colors.get(s, "gray") for s in subset["sector"]]
    )
    ax.set_title(str(year))
    ax.set_xlabel("Sector")
    ax.set_ylabel("Total Service User Count")
    ax.set_xticks(range(len(subset)))
    ax.set_xticklabels(subset["sector"], rotation=45, ha='right')
    ax.yaxis.set_major_formatter(FuncFormatter(lambda x, _: f"{int(x):,}"))

# Remove unused subplots
for j in range(i + 1, len(axes)):
    fig.delaxes(axes[j])

# Add shared legend
handles = [plt.Rectangle((0, 0), 1, 1, color=c) for c in sector_colors.values()]
labels = list(sector_colors.keys())
fig.legend(handles, labels, title="Sector", loc="center right")

plt.show()
```
